Adaptive group-regularized logistic elastic net regression

### 2019-04-01 docs/manuscript_biostatistics.Rnw
added all suggestions

\documentclass[oupdraft]{bio}
\usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
\usepackage{url,bm,algorithm,algpseudocode,mathtools,bbm,pgfplotstable}

@@

	Recently, \cite{zhang_novel_2018} introduced VBVS, a method similar to ours in
	spirit. Nevertheless, one clear advantage of our method over VBVS is the
	possibility of adaptive inclusion of external information on the features to
	aid in prediction and feature selection. Bayesian versions of support vector 
	machines have been used in classification problems as well 
	\cite[]{chakraborty_bayesian_2011}, but these methods also lack 
	adaptive inclusion of external information on the features.

	In line with the above, our proposed method is (a) subjective: the 
	hyper-parameters are estimated from the data and not chosen on objective 
	grounds, (b) adaptive: prior information is automatically and subjectively 
	weighed with respect to its informativeness, (c) fast compared to full 
	Bayesian analysis or cross-validation, and (d) requires minimal input by the 
	user: only the data and grouping of the features is required. We emphasize that 
	our method is different from recent developments in grouped penalized 
	regression such as \cite{dondelinger_joint_nodate}, where instead of the 
	features, the observations are partitioned.

	The rest of the paper is structured as follows. We introduce the Bayesian 
	logistic generalised elastic net model in Section
	\ref{sec:model}. In Section \ref{sec:estimation} we derive a variational
	approximation to this model and use this novel approximation in the empirical
	Bayes estimation of multiple, group-specific penalty parameters. In Section
	\ref{sec:simulations} we compare the method in a simulation study. In Sections
	\ref{sec:colorectal} and \ref{sec:lymphnode}, we demonstrate the approach in 
	two data sets. In Section \ref{sec:discussion} 
	we conclude by both discussing some differences and
	parallels between the proposed method, termed \texttt{gren}, and other methods
	in the literature.

@@

	\section{Simulations}\label{sec:simulations}
	To assess classification and variable selection performance we conducted a 
	simulation study in which we compared \texttt{gren} to the regular elastic 
	net and ridge models, \texttt{GRridge} \cite[]{van_de_wiel_better_2016}, 
	sparse group lasso (\texttt{SGL}) \cite[]{simon_sparse-group_2013}, 
	composite mimimax concave penalty (\texttt{cMCP}) 
	\cite[]{breheny_penalized_2009}, and the group exponential lasso 
	(\texttt{gel}) \cite[]{breheny_group_2015}. \texttt{GRridge} is similar to 
	\texttt{gren} in the sense that it estimates group-specific penalty 
	multipliers. The two main differences with \texttt{gren} are (i) the absence 
	of an $L_1$-norm penalty and (ii) the estimation procedure. The other methods 
	are extensions of the group lasso and not adaptive on the group level. 
	However, in contrast to the original group lasso,
	they select single features, instead of just whole groups.

	For the regular elastic net, \texttt{gren}, \texttt{SGL}, \texttt{cMCP}, 
	and \texttt{gel}, we fix $\alpha \in \{0.05, 0.5, 0.95 \}$. In \texttt{gren}
	and the elastic net, $\alpha$ determines the proportion of $L_1$-norm penalty,
	such that the first setting closely resembles the ridge setting, 
	where $\alpha=0$, while the third setting is similar to the lasso with 
	$\alpha=1$. In \texttt{SGL}, \texttt{cMCP}, and \texttt{gel}, $\alpha$
	determines the ratio of group-level penalization ($\alpha=0$) to 
	feature level ($\alpha=1$) penalization. For the methods that do 
	automatic feature selection, we estimate models for a range of model sizes:
	$\hat{p} \in \{2, 4, 8, 16, 32, 64, 128, 256\}$.

	We simulated data according to five different scenarios:
	\begin{enumerate}
		\item differential signal between the groups and uniformly distributed model
		parameters,
		\item a large number of small groups of features,
		\item no differential signal between the groups, but strong correlations 
		within groups of features,
		\item differential signal between the groups and heavy-tailed distributed 
		model parameters, and
		\item a very sparse setting with no signal in some of the groups.
	\end{enumerate}
	Full descriptions of the scenarios are given in Section 
	\ref{sm-sec:simulations} of 
	the SM. In the first scenario, the distribution of model parameters is far
	from the assumed ridge/elastic net/lasso distributions, so we expect
	all methods to underperform. In the second scenario, which is similar to
	the simulations presented in the original group lasso paper 
	\cite[]{yuan_model_2006}, we expect \texttt{gren}
	and \texttt{GRridge} to overfit due to the large number of hyperparameters to 
	estimate. The group lasso extensions are expected to perform better: they 
	estimate just one penalty parameter, so are less sensitive to overfitting.
	In scenario three, we expect all methods to underperform
	due to the high correlations between the features. The fourth 
	scenario is expected to yield a better \texttt{gren} performance, since
	\texttt{gren} assumes a heavier-tailed distribution and is designed to 
	detect differential signal between
	the groups. Scenario five is expected to yield a bad performance for all 
	methods, because of the sparse truth and lack of differential 
	signal.

	In all scenarios we created a training set of $n=100$ samples to estimate the 
	models and a test set of $n_{\text{test}}=1000$ samples to compute 
	the performance measures. The features $\x_i$, $i=1, \dots, n$, are taken 
	from a $p$-dimensional zero-centred Gaussian distribution, where we set 
	$p=900$ in scenario 2 and $p=1000$ in the other scenarios. We varied the 
	covariance matrix of the features and simulation of 
	the model parameters $\beta_j$, $j=1, \dots, p$ between the
	scenarios (see SM Section \ref{sm-sec:simulations} for the exact 
	parametrisations). We 
	simulated the outcomes $y_i$, $i=1, \dots, n$, from a binary logistic model: 
	$y_i \sim \mathcal{B}(1, \expit (\x_i \tr \bbeta))$. To mitigate the influence
	of random variation, we repeated each scenario 100 times 
	and report predictive performance measures 
	area under the receiver operator curve (AUC) and Brier skill score (BSS) in 
	Figures \ref{fig:lines_simulations_res1_auc_briers}--[! figure refs here]. 
	Additional results and performance measures are presented in Section 
	\ref{sm-sec:simulations} of the SM.

	- description of results

@@

  We compared the performance of \texttt{gren} to ridge, \texttt{GRridge}, 
  elastic net, \texttt{SGL}, \texttt{cMCP}, and \texttt{gel}. Of the latter 
  three methods, we only present \texttt{cMCP} here; the results for 
  \texttt{SGL} and \texttt{gel} are presented in the SM Section 
  [! section ref here]. For the methods with a
  tuning parameter $\alpha$, we picked the $\alpha$('s) that gave the best
  performance and refer
  the reader to SM Section \ref{sm-sec:colorectal} for comparisons with 
  different $\alpha$'s.

@@

  The penalty multipliers estimated by 
  \texttt{gren} and \texttt{GRridge} are according to expectation: the highly
  expressed group receives the lowest penalty, followed by the medium expressed
  group, while the non-expressed group receives the strongest penalty 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res1_auc}a). 
  \texttt{gren} seems to outperform the other methods for a range of model 
  sizes in terms of AUC 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res1_auc}b). 
  This is especially pronounced for the larger models. The performance of
  \texttt{cMCP} is somewhat unstable. This unstable estimation is an issue for   
  all group lasso extensions that we investigated.
  (SM Section \ref{sm-sec:colorectal}). Overall, it seems as though the 
  inclusion of the extra information on the 
  features benefited predictive performance: both \texttt{GRridge} and 
  \texttt{gren} outperform their respective non-group-regularized versions, 
  albeit it only slightly for \texttt{GRridge}.

  \section{Application to RNAseq in lymph node cancer}\label{sec:lymphnode}
  The aim of our second example study was to predict lymph node metastasis in
  oral cancer patients using sequenced RNAs from TCGA 
  \cite[]{the_cancer_genome_atlas_network_comprehensive_2015}. The features are 
  3096 transformed and normalised TCGA RNASeqv2 profiles of head-and-neck 
  squamous cell carcinomas on 133 patients with HPV-negative tumours in the oral
  cavity. Of these patients, 76 suffered from lymph node metastasis, while 57 
  did not. For a thorough introduction of these data, see 
  \cite{te_beest_improved_2017}.

  We considered two sources of external feature information: (a) the 
  cis-correlations between the RNASeqv2 data and TCGA DNA copy numbers on 
  the same patients, quantified by Kendall's $\tau$ and binned into five groups.
  In addition, we used (b) $p$-values from a microarray data set
  described in \cite{mes_prognostic_2017}, again binned into five groups. 
  We expected features 
  with a large positive Kendall's $\tau$ to be more important in 
  metastasis prediction \cite[]{masayesva_gene_2004}. Likewise, we expect 
  features with low external $p$-values to be more important.

  We compared \texttt{gren} to the same methods as in Section 
  \ref{sec:colorectal}. However, since the feature information
  consists of two partitions with overlapping groups, we used an extension of
  \texttt{cMCP} that allows for
  overlapping groups \cite[]{zeng_overlapping_2016}.
  We estimated AUC for the methods on an independent validation set of 97 
  patients \cite[]{mes_prognostic_2017}. The features 
  in this validation set were
  microarray expression levels, normalised to account
  for a difference in scale compared to the RNAseq training data. We present the
  results on this validation set, together with the estimated penalty 
  multipliers for \texttt{gren} and \texttt{GRridge}, in Figure 
  \ref{fig:barplot_lines_rnaseq_lymph_node_metastasis_res1_auc}. Additional
  comparisons and performance measures are presented in the SM Section 
  \ref{sm-sec:lymphnode}.

>>>>

\documentclass[oupdraft]{bio}
\usepackage{xr-hyper}
% \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, 
% citecolor=citecolor]{hyperref}
\usepackage{url,bm,algorithm,algpseudocode,mathtools,bbm,pgfplotstable,
enumerate}

\externaldocument[sm-]{supplement}

@@

	In line with the above, our proposed method is (a) subjective: the 
	hyper-parameters are estimated from the data and not chosen on objective 
	grounds, (b) adaptive: prior information is automatically and subjectively 
	weighed with respect to its informativeness, (c) fast compared to full 
	Bayesian analysis or cross-validation, and (d) requires minimal input by the 
	user: only the data and grouping of the features is required. We emphasize 
	that our method is different from recent developments in grouped penalized 
	regression such as \cite{dondelinger_joint_2018}, where instead of the 
	features, the observations are partitioned.

	Recently, \cite{zhang_novel_2019} introduced VBVS, a method similar to ours in
	spirit. Nevertheless, one clear advantage of our method over VBVS is the
	possibility of adaptive inclusion of external information on the features to
	aid in prediction and feature selection. Bayesian versions of support vector 
	machines have been used in classification problems as well 
	\cite[]{chakraborty_bayesian_2011}, but these methods also lack 
	adaptive inclusion of external information on the features.

	The rest of the paper is structured as follows. We introduce the Bayesian 
	logistic generalised elastic net model in Section
	\ref{sec:model}. In Section \ref{sec:estimation} we derive a variational
	approximation to this model and use this novel approximation in the empirical
	Bayes estimation of multiple, group-specific penalty parameters. In Section
	\ref{sec:simulations} we compare the method in a simulation study. In Sections
	\ref{sec:colorectal} and \ref{sec:oral}, we demonstrate the approach in 
	two data sets. In Section \ref{sec:discussion} 
	we conclude by both discussing some differences and
	parallels between the proposed method, termed \texttt{gren}, and other methods
	in the literature.

@@

	\section{Simulations}\label{sec:simulations}
	To assess classification and variable selection performance we conducted a 
	simulation study in which we compared \texttt{gren} to the regular elastic 
	net and ridge models, \texttt{GRridge} \cite[]{van_de_wiel_better_2016},     
	composite mimimax concave penalty (\texttt{cMCP}) 
	\cite[]{breheny_penalized_2009}, and the group exponential lasso 
	(\texttt{gel}) \cite[]{breheny_group_2015}. \texttt{GRridge} is similar to 
	\texttt{gren} in the sense that it estimates group-specific penalty 
	multipliers. The two main differences with \texttt{gren} are (i) the absence 
	of an $L_1$-norm penalty and (ii) the estimation procedure. The other methods 
	are extensions of the group lasso and not adaptive on the group level. 
	However, in contrast to the original group lasso,
	they select single features, instead of complete groups.

	For the regular elastic net, \texttt{gren}, \texttt{cMCP}, 
	and \texttt{gel}, we fix $\alpha \in \{0.05, 0.5, 0.95 \}$. In \texttt{gren}
	and the elastic net, $\alpha$ determines the proportion of $L_1$-norm penalty,
	such that the first setting closely resembles the ridge setting, 
	where $\alpha=0$, while the third setting is similar to the lasso with 
	$\alpha=1$. In \texttt{cMCP} and \texttt{gel}, $\alpha$
	determines the ratio of group-level penalization ($\alpha=0$) to 
	feature level ($\alpha=1$) penalization. For the methods that do 
	automatic feature selection, we estimate models for a range of model sizes:
	$\hat{p} \in \{2, 4, 8, 16, 32, 64, 128, 256\}$.

	We simulated data according to five different scenarios:
	\begin{enumerate}[(i)]
		\item \label{it:scenario1} differential signal between the groups and uniformly
		distributed model
		parameters,
		\item \label{it:scenario2} a large number of small groups of features,
		\item \label{it:scenario3} no differential signal between the groups, but 
		strong correlations 
		within groups of features,
		\item \label{it:scenario4} differential signal between the groups and 
		heavy-tailed distributed 
		model parameters, and
		\item \label{it:scenario5}a very sparse setting with no signal in some of 
		the groups.
	\end{enumerate}
	Full descriptions of the scenarios are given in Section 
	\ref{sm-sec:simulations} of 
	the SM. In Scenario (\ref{it:scenario1}), the distribution of model 
	parameters is far
	from the assumed ridge/elastic net/lasso distributions, so we expect
	all methods to underperform. In Scenario (\ref{it:scenario2}), 
	which is similar to
	the simulations presented in the original group lasso paper 
	\cite[]{yuan_model_2006}, we expect \texttt{gren}
	and \texttt{GRridge} to overfit due to the large number of hyperparameters to 
	estimate. The group lasso extensions are expected to perform better: they 
	estimate just one penalty parameter, so are less sensitive to overfitting.
	In Scenario (\ref{it:scenario3}), we expect all methods to underperform
	due to the high correlations between the features. The
	Scenario (\ref{it:scenario4}) is expected to yield a better \texttt{gren}
	performance, since
	\texttt{gren} assumes a heavier-tailed distribution and is designed to 
	detect differential signal between
	the groups. Scenario (\ref{it:scenario5}) is expected to yield a bad 
	performance for all 
	methods, because of the sparse truth and lack of differential 
	signal.

	In all scenarios we created a training set of $n=100$ samples to estimate the 
	models and a test set of $n_{\text{test}}=1000$ samples to compute 
	the performance measures. The features $\x_i$, $i=1, \dots, n$, are taken 
	from a $p$-dimensional zero-centred Gaussian distribution, where $p$ varies 
	from $p=900$ to $p=1000$ over the Scenarios. We varied the 
	covariance matrix of the features and simulation of 
	the model parameters $\beta_j$, $j=1, \dots, p$ between the
  Scenarios (see SM Section \ref{sm-sec:simulations} for the exact 
	parametrisations). We 
	simulated the outcomes $y_i$, $i=1, \dots, n$, from a binary logistic model: 
	$y_i \sim \mathcal{B}(1, \expit (\x_i \tr \bbeta))$. To mitigate the influence
	of random variation, we repeated each scenario 100 times 
	and report predictive performance measures 
	area under the receiver operator curve (AUC) and Brier skill score (BSS) in 
	Figures \ref{fig:lines_simulations_res1_auc_briers}--[! figure refs here]. 
	Additional results and performance measures are presented in Section 
	\ref{sm-sec:simulations} of the SM.

	In Scenario (\ref{it:scenario1})
	\texttt{gren} and \texttt{GRridge} don't suffer much from the wrong 
	distributional assumptions as they outperform the other methods.

	In Scenario (\ref{it:scenario3}), \texttt{gren} and to a lesses extent the
	regular elastic net suffer more from the higher correlations. Here,
	the regular elastic net outperforms \texttt{gren}, which is an indication
	that the estimation of the penalty parameters is impaired by high
	correlation between features. Scenario (\ref{it:scenario4}) follows the 
	expected pattern, with the group-regularized methods (\texttt{GRridge} and
	\texttt{gren}) outperforming their respective non group-regularized 
	counterparts, as well as the sparse group lasso extensions.
	Surprisingly, in Scenario (\ref{it:scenario5}), \texttt{gren} outperforms all
	other methods, which is an indication that \texttt{gren} is able to pick up 
	very sparse differential signals between the groups.

@@

  We compared the performance of \texttt{gren} to ridge, \texttt{GRridge}, 
  elastic net, \texttt{SGL}, \texttt{cMCP}, and \texttt{gel}. Of the latter 
  three methods, we only present \texttt{cMCP} here; the results for 
  another extension of the group lasso, the sparse
  group lasso by \cite{simon_sparse-group_2013}
  (\texttt{SGL}), and \texttt{gel} are presented in the SM Section 
  [! section ref here]. For the methods with a
  tuning parameter $\alpha$, we picked the $\alpha$ 
  (two $\alpha$'s for \texttt{gren}) that gave the best
  performance and refer
  the reader to SM Section \ref{sm-sec:colorectal} for comparisons with 
  different $\alpha$'s.

@@

  The penalty multipliers estimated by 
  \texttt{gren} and \texttt{GRridge} are according to expectation: the highly
  expressed group receives the lowest penalty, followed by the medium expressed
  group, while the non-expressed group receives the strongest penalty 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res1_auc}a). 
  \texttt{gren} seems to outperform the other methods for a range of model 
  sizes in terms of AUC 
  (Figure \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res1_auc}b). 
  This is especially pronounced for the larger models. The performance of
  \texttt{cMCP} is somewhat unstable. This unstable estimation is an issue for   
  all group lasso extensions that we investigated.
  (SM Section \ref{sm-sec:colorectal}). Overall, the 
  inclusion of the extra information on the 
  features benefited predictive performance: both \texttt{GRridge} and 
  \texttt{gren} outperform their respective non-group-regularized versions, 
  albeit it only slightly for \texttt{GRridge}.

  \section{Application to RNAseq in oral cancer}\label{sec:oral}
  The aim of our second example study is to predict lymph node metastasis in
  oral cancer patients using sequenced RNAs from TCGA 
  \cite[]{the_cancer_genome_atlas_network_comprehensive_2015}. The features are 
  3096 transformed and normalised TCGA RNASeqv2 expression values of 3096 genes 
  for 133 HPV-negative oral tumours. Of the corresponding patients, 76 suffered 
  from lymph node metastasis, while 57 
  did not. For a thorough introduction of these data, see 
  \cite{te_beest_improved_2017}.

  We considered two sources of external feature information: (a) the 
  cis-correlations between the RNASeqv2 data and TCGA DNA copy numbers on 
  the same patients, quantified by Kendall's $\tau$ and binned into five groups.
  In addition, we used (b) $p$-values from an independent microarray data set
  described in \cite{mes_prognostic_2017}, again binned into five groups. 
  We expected features 
  with a large positive Kendall's $\tau$ to be more important in 
  metastasis prediction \cite[]{masayesva_gene_2004}. Likewise, we expect 
  features with low external $p$-values to be more important.

  We compared \texttt{gren} to the same methods as in Section 
  \ref{sec:colorectal}. However, since the feature information
  consists of two partitions with overlapping groups, we used an extension of
  \texttt{cMCP} that allows for
  overlapping groups \cite[]{zeng_overlapping_2016}.
  We estimated AUC for the methods on an independent validation set of 97 
  patients \cite[]{mes_prognostic_2017}. The features 
  in this validation set were
  microarray expression levels, normalised to account
  for a difference in scale compared to the RNAseq training data. We present the
  results on this validation set, together with the estimated penalty 
  multipliers for \texttt{gren} and \texttt{GRridge}, in Figure 
  \ref{fig:barplot_lines_rnaseq_lymph_node_metastasis_res1_auc}. Additional
  comparisons and performance measures are presented in the SM Section 
  \ref{sm-sec:oral}.
